---
title: dplyr Overview Part 2
author: David Pritchard
date: April 7, 2020
output:
  revealjs::revealjs_presentation:
    theme: serif
    center: false
---


```{r, echo = FALSE}
options(tibble.print_max = 6)
options(tibble.print_min = 6)

knitr::opts_chunk$set(collapse = TRUE)

# def.chunk.hook  <- knitr::knit_hooks$get("chunk")
# knitr::knit_hooks$set(chunk = function(x, options) {
#   x <- def.chunk.hook(x, options)
#   ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
# })
# knitr::opts_chunk$set(size = "small")
```


# Motivation

## Historical perspective

Metaprogramming

  * The S language bears a strong similarity to Scheme[^scheme] (with some
    notable exceptions).
  * One particular language feature that is fundamental to Lisp is a form of
    metaprogramming in which programs can create or manipulate other programs
    at runtime.  <!-- Scott page 562 -->

[^scheme]: A Lisp dialect created by Guy Steele and Gerald Jay Sussman.




## Historical perspective (cont.)

Metaprogramming


  * One form of metaprogramming that is available in many languages is the
    ability to `eval` text strings.
  * In Lisp there is no syntactical differentiation between code and data, so
    metaprogramming involves direct manipulating of the program.
  * S provides an interface to manipulate the data structure that programs are
    stored in (called the AST) since S3 (S version 3) in order to provide
    essentially the same functionality.




## Abstract syntax trees

Programs are naturally represented by the interpreter as a tree-like data
structure.

  * The read portion of the read-eval-print loop constructs the AST.
  * The R interpreter (in contrast to S) is implemented as a lisp.

    * R's infix operators are translated to Polish notation (i.e. operators
      precede their arguments).
    * So, for example, `x + y` is translated internally to `"+"(x, y)`




## Abstract syntax tree example

Consider the AST resulting from the expression `1 + 2 * 3`.  The AST has the
following form:

  * The result of evaluating `+` on `1` and ...
    * The result of evaluating `*` on `2` and `3`

```{r}
library(lobstr)
ast(1 + 2 * 3)
```

You can construct a data structure representing an AST by using e.g. `quote(1 +
2 * 3)`.




## Abstract syntax tree components

There are four types of data structures that can be found in an AST, which are
collectively known as **expressions**:

  * **Constant:** either `NULL` or a length-1 atomic vector.
  * **Symbol:** the name of an object like e.g. `x`.  (These are synonymously
    known as **name**s.)
  * **Call**: A function call.
  * **Pairlist**: a list-like object most often used to store a function's
    formal arguments.




## Language objects

R has a notion of what it calls language objects:

  * consists of calls, symbols, and expression vectors.

An expression vector is a list of R expressions.

```{r}
a <- expression(x + y, z)
b <- parse(text = "x + y; z")  # equivalent up to the attributes
is.expression(a)
is.expression(b)
a
b
```




## Evaluating expressions

You can use `eval` to evaluate an expression or expression vector.
```{r}
eval(parse(text = "x <- 2; x * x"))
eval(1)
eval(quote(mean(1L:11L)))
eval(quote(x))
```

It is interesting to note that in most languages, `eval` takes a string as
input.

  * Since the elements of the AST are exposed to the user, `eval` operates on
    them directly.




## Examples of common AST elements

```{r}
identical(quote(1), 1)  # by definition, constants are self-evaluating
is.language(1)

b <- quote(x)
is.language(b); is.symbol(b)

c <- quote(1 + 2 * 3)
is.language(c); is.call(c)
```




## Language objects (cont.)

Some notes about language objects:

  * Hadley Wickham and Lionel Henry use the term _expression_ to mean any of the
    data structures found in an AST, and refer to the `expression` object as
    _expression vectors_.  **We will follow suit.**
  * Hadley suggests using lists of calls instead of expression vectors in most
    cases.




## Manipulating calls

You can access and modify elements of a call using a list-like interface.

```{r, collapse = TRUE}
z <- quote(1 + 2 * 3)

z[[1L]]
is.symbol(z[[1L]])  # synonymous with `is.name`

z[[2L]]
is.numeric(z[[2L]])

z[[3L]]
is.call(z[[3L]])
```




## Manipulating calls (cont.)


```{r, collapse = TRUE}
z
ast(1 + 2 * 3)

length(z)
z[[1L]] <- as.name("-")  # synonymous with `as.symbol`
z[[2L]] <- quote(6 / 2)
z
```




## Example parsing input

```{r}
xdist = "x = rnorm(1,y,1)
y = rnorm(1,z,1)
z = rnorm(1,5,1)"
```




## `parse_heir_minilang` definition

```r
parse_hier_minilang <- function(hierarchical) {
  comb_hier <- function(outerdist, innerdist) {
    out <- paste(
      strsplit(outerdist[2L], innerdist[1L])[[1L]],
      collapse = innerdist[2L]
    )
    c(outerdist[[1L]], out)
  }
  distexpr <- strsplit(
    strsplit(hierarchical, "\n")[[1L]],
    " *= *"
  )
  distcomb <- Reduce(comb_hier, distexpr)
  distcomb[[2L]]
}
heir_dist_expr <- parse_hier_minilang(xdist)
n_sims <- replicate(1000L, eval(parse(text = heir_dist_expr)))
```




# Promises and lazy evaluation

## Lazy evaluation in R

The S language stipulates that the arguments of a function call are not
evaluated until they are needed.

  * A form of _lazy evaluation_.
  * Implemented in R by assigning each formal argument name a special object of
    type _promise_.




## Promises

The _promise_ type has the following characteristics.

  * Contains three elements:
    * The expression provided as input to the argument in the function call.
    * The environment (i.e. the scope) in which to evaluate the expression.
    * The value obtained by evaluating the expression (initially empty).
  * The value of the promise is calculated the first time it is needed, and not
    before.  After calculation the value is stored.
  * Subsequent uses of the variable use the stored value.




## Extracting the promise expression

The `substitute` function creates an expression in a manner similar to `quote`.

  * However, `substitute` performs the added step of replacing symbols that
    it finds in the expression that exist in the specified environment.
    * The environment is specified through the formal argument `env` (default is
      the current environment).
    * If a symbol in the expression has a binding in the specified environment,
      then it replaces the symbol with the value bound to the symbol in the
      environment.




## Extracting the promise expression (cont.)


```{r}
print_input <- function(x) {
  substitute(x)
}
print_input(1 + 2 * 3)
```




<!-- promises page 457 Software for Data Analysis ----------------------------->

<!-- R implements the grammer of S, with a few extensions --------------------->

<!-- http://www.sumsar.net/blog/2014/11/tidbits-from-books-that-defined-s-and-r/ -->

<!-- https://www.stat.auckland.ac.nz/~ihaka/downloads/Interface98.pdf -->
