---
title: dplyr Overview Part 2
author: David Pritchard
date: April 7, 2020
output:
  revealjs::revealjs_presentation:
    theme: serif
    # theme: simple
    center: false
---

<style>
  .reveal section p {
    display: inline-block;
    font-size: 0.7em;
    line-height: 1.2em;
    vertical-align: top;
    text-align: left;
  }
  .reveal h1 {
    font-size: 1.8em;
    color: darkblue;
  }
  .reveal h2 {
    font-size: 1.2em;
  }
  .reveal ul {
    font-size: 0.7em;
  }
  .reveal ul ul {
    font-size: 0.9em;
    <!-- line-height: 350%; -->
    <!-- margin-bottom: 45px; -->
  }
  .reveal li {
    line-height: 100%;
    margin-bottom: 15px;
    margin-top: 00px;
  }
  .reveal li li {
    margin-top: 10px;
    <!-- margin-bottom: 10px; -->
  }
</style>


```{r, echo = FALSE}
options(tibble.print_max = 6)
options(tibble.print_min = 6)

knitr::opts_chunk$set(collapse = TRUE)
suppressWarnings(suppressMessages(library(dplyr)))
```


# Motivation

## Motivating examples

The R package dplyr is arguably one of the most valuable tools in the R
ecosystem.  dplyr makes use of a number of facilities in the R programming
language to provide a high level of ergonomics.  Consider the following
examples.

```{r, results="hide"}
library(dplyr)
mutate(mtcars, y = 0.45 * wt, z = 1000 * y)
select(iris, -starts_with("Petal"))
```

In the first example, `wt` refers to a variable name in `mtcars`, while `z`
refers to a variable `y` that doesn't exist at call time.  In the second
example, the `select` function provides the user with a domain-specific language
that is used to remove any variables starting with `"Petal"`.




## Tidy evaluation

The niceties demonstrated in the previous slide do come at a cost: they sometime
require some extra techniques when using them, as anyone who has programmed with
dplyr for some time will have discovered.

  * The custom evaluation used by dplyr and other libraries is powered by the
    rlang library and which is referred to by its authors as _tidy evaluation_.
  * In the upcoming slides, we will aim to cover the set of concepts of
    foundational concepts upon which rlang is constructed.




## Historical perspective

Metaprogramming

  * The S language bears a strong similarity to Scheme[^scheme] (with some
    notable exceptions).  <!-- paraphrased from Ihaka R A Language for Data
    Analysis and Graphics -->
  * One particular language feature that is fundamental to Lisp is a form of
    metaprogramming in which programs can create or manipulate other programs
    at runtime.  <!-- Scott page 562 -->
  * In Lisp there is no syntactical differentiation between code and
    data[^homoiconicity], so metaprogramming involves direct manipulating of the
    program.  S provides an interface to manipulate the data structure (more
    upcoming on this) that statements are stored in order to provide similar
    functionality.  <!--
    http://www.sumsar.net/blog/2014/11/tidbits-from-books-that-defined-s-and-r/
    -->

[^scheme]: A Lisp dialect created by Guy Steele and Gerald Jay Sussman.

[^homoiconicity]: This property is sometimes referred to as _homoiconicity_,
however there is some disagreement about the usage of the term, see
e.g. https://www.expressionsofchange.org/dont-say-homoiconic/.




<!-- ## Historical perspective (cont.) -->

<!-- Metaprogramming -->

<!--   * One familiar form of metaprogramming that is available in many languages is -->
<!--     the ability to `eval` text strings. -->
<!--   * In Lisp there is no syntactical differentiation between code and -->
<!--     data[^homoiconicity], so metaprogramming involves direct manipulating of the -->
<!--     program.  S provides an interface to manipulate the data structure (more -->
<!--     upcoming on this) that statements are stored in order to provide similar -->
<!--     functionality.  <\!-- -->
<!--     http://www.sumsar.net/blog/2014/11/tidbits-from-books-that-defined-s-and-r/ -->
<!--     -\-> -->

<!-- [^homoiconicity]: This property is sometimes referred to as _homoiconicity_, -->
<!-- however there is some disagreement about the usage of the term, see -->
<!-- e.g. https://www.expressionsofchange.org/dont-say-homoiconic/. -->




## Abstract syntax trees

Expressions in programming languages are naturally represented by the
interpreter as a tree-like data structure called an **abstract syntax tree**
(AST).

  * The read portion of the read-eval-print loop constructs the AST.
  * The R interpreter (in contrast to S) is implemented as a Lisp.
    * R's infix operators are translated to Polish notation (i.e. operators
      precede their arguments, and consequently is also known as prefix
      notation).
    * So, for example, `x + y` is translated internally to `"+"(x, y)`
    * Even structures like if-else if-else and for loops that are represented as
      syntax in many languages are mapped to prefix functions.




## Abstract syntax tree example

Consider the AST constructed from the expression `1 + 2 * 3`.  The AST has the
following form:

  * The result of evaluating `+` on `1` and
    * The result of evaluating `*` on `2` and `3`

The `lobstr::ast` function provides a visual representation of an AST
constructed from an R statement.

```{r}
library(lobstr)
ast(1 + 2 * 3)
```




## Why was S so closely modeled after Scheme?

In 1986, the new version of S titled S3, in the words of John Chambers, featured

> dynamically generated, self-describing objects and a strong (though not
> strict) adherence to the ideas of functional programming.

Chambers further writes that

> The goal of functional programming has much in common with our _Prime
> Directive_: to provide software that is understandable and as demonstrably
> correct as possible.

Given a lack of actual information regarding this topic, I delve into the realm
of speculation to propose that Scheme was viewed as a paragon of a functional
programming language, and is one that has thankfully stood up to the test of
time.




## Abstract syntax tree components

There are four types of data structures that can be found in an AST in R, which
are collectively referred to as _expressions_ by Wickham, Henry, etc.: <!--
Section 18.3, Advanced R -->

  * **Call**: A function call.
  * **Constant:** either `NULL` or a length-1 atomic vector.
  * **Symbol:** the name of an object like e.g. `x`.  (These are synonymously
    known in R as **name**s.)
  * **Pairlist**: a list-like object most often used to store a function's
    formal arguments.




## A note about terminology:

Hadley Wickham and Lionel Henry have co-opted the term _expression_ mean any of
the data structures found in an AST, however the R Language Definition does not
use this terminology.

  * The term _expression_ is used by Wickham and Henry to mean any of the data
    structures found in an AST
  * R has an `expression` data type, which Wickham and Henry refer to as
    _expression vectors_.
  * In the general computer science literature, an expression is a combination
    of one or more constants, variables, operators, and functions that the
    programming language interprets.  The R Language definition refers to these
    as _statements_ to disambiguate from expression vectors.

In this document, we will follow the convention proposed by Wickham and Henry.




## Language objects

R has also has related concepts of language objects and language data types, so
let's get these terminologies straight as well.

  * A language object is an umbrella term referring to an object that is either
    a call, expression, or symbol.
    * `is.language` returns `TRUE` for any one of these three data types, and
      `FALSE` otherwise.
    * `typeof` may return `language` (see below), `expression`, or `symbol`,
      depending on the object type.
  * Confusingly, R also has a language data type which is the data type used to
    represent calls.
    * `is.language` returns `TRUE`, `is.call` returns `TRUE`, `typeof` returns
      `language`, and `mode` returns `call`.

If you think this is unfortunate vocabulary, then you are not alone.




<!----------------------------------------------------------------------------->

# Capturing and evaluating expressions

## Typical expression evaluation

One of the tasks that is performed by an interpreter (indeed, the interpreter's
raison d'être) such as R is to evaluate expressions (i.e. the 'eval' stage of
the read-eval-print loop).

  * Evaluating an expression produces a value.
  * Expressions may consist of sub-expressions (e.g. `f(1, g(7, 2))`).
  * Sub-expressions are recursively evaluated until they cannot be simplified
    any further (at which point the expression is said to be in _normal form_).




## Quoting code

In order to be able to obtain a data structure representing an unevaluated
expression or sub-expression, Lisp-like languages such as R provide a mechanism
called _quotation_, which is a function that simply returns its input unmolested
by the evaluator.  R provides several interfaces for quotation.

  * The `quote` function is perhaps the most fundamental expression of
    quotation in R, but as we will see there are many others.

```{r}
quote(1 + 2 * 3)
```

As an aside, note that it is common to refer to "quoting" a section of source
code to mean placing single or double quotation marks around that code, but this
is an entirely separate meaning of the word "quoting."




## Manipulating expressions

You can access and modify expressions using a list-like interface.

```{r, collapse = TRUE}
lobstr::ast(f(1, g(7, 2)))
z <- quote(f(1, g(7, 2)))
z

z[[3L]][[3L]] <- quote(h(99))
z
```




## Expression vectors

Expression vectors are a data type providing an ordered collection of
expressions.  The functions `expression` and `parse` each return expression
vectors and can be used to capture expressions.

  * The `expression` function performs quotation on each of a sequence of
    comma-separated expressions.
  * The `parse` function parses a sequence of statements either from a source
    file or a text string.

```{r}
# `a` and `b` are equivalent up to the attributes
a <- expression(x + y, z)
b <- parse(text = "x + y; z")
all.equal(a, b, check.attributes = FALSE)
typeof(a)
a
```




## Lists of expressions with alist

TODO




## Constructing text from an expression

TODO: deparse




## Quotation in rlang

The rlang library provides several alternatives for quotation.

  * `rlang::expr` is a replacement for `quote`.
  * `rlang::exprs` is a replacement for `alist` (and serves a similar purpose to
    `expression`).
  * `rlang::parse_expr` and `rlang::parse_exprs` are similar to `parse(text =
    "somestring")`, (the former rlang variant returns an expression, and the
    latter returns a list of expressions).

```{r}
identical(quote(22 * y), rlang::expr(22 * y))
z = rlang::exprs(a = 1, b = x, c = f("one", 2))
str(z)
zz = list(a = quote(1), b = quote(x), c = quote(f("one", 2)))
identical(z, zz)
```




## Evaluating expressions

You can use `eval` to evaluate an expression or expression vector.
```{r}
eval(parse(text = "x <- 2; x * x"))
eval(1)
eval(quote(mean(seq(1, 101))))
eval(quote(x))
```

  * It is interesting to note that in most languages, `eval` takes a string as
    input.
  * Since the elements of the AST are exposed to the user in R, `eval` operates
    on them directly.




<!----------------------------------------------------------------------------->

# Expression elements

## Expression elements: constants

As previously mentioned, constants are either `NULL` or a length-1 atomic vector
such as `TRUE`, `1L`, `2.5`, `1i`, or `"mystring"`.

  * These types each have a special read syntax that the parser recognizes,
    creating an object with the specific type.
  * The defining property of constants in general is that they are _self
    evaluating_, meaning that the result of evaluating the constant is the
    constant itself.

```{r}
# expression vectors can be treated as a list-like object
sapply(expression(TRUE, 1L, 2.5, 1i, "mystring"), typeof)
sapply(expression(TRUE, 1L, 2.5, 1i, "mystring"), rlang::is_syntactic_literal)
identical(2.5, quote(2.5))
```




## Expression elements: symbols

A symbol represents the name of an object, such as `x` in the expression `x +
1`.

  * When the R evaluator evaluates a symbol, it replaces it with the value that
    is bound to the appropriate variable of the same name
  * The variable is selected according to the lexical scoping rules.

```{r}
quote(x)
c(is.symbol(x), rlang::is_symbol(x))
typeof(quote(x))
identical(quote(x), as.symbol("x"))
identical("x", as.character(quote(x)))
```




## Expression elements: pairlists

Pairlists are most often encountered as the data structure representing the formal
arguments to a function.

  * Pairlists are implemented using a linked list, but can be treated just like
    a regular list.

```{r}
lobstr::ast(function(x = 2) 22)
f <- quote(function(x = 2) 22)
f[[2L]]
c(is.pairlist(f[[2L]]), rlang::is_pairlist(f[[2L]]))
typeof(f[[2L]])
```




## Expression elements: calls

A call is a unevaluated function call.

```{r}
lobstr::ast(f(1, g(7, 2)))
x = quote(f(1, g(7, 2)))
is.language(x)
c(is.call(x), rlang::is_call(x))
typeof(x)
mode(x)
```




## Constructing calls

TODO




## Formulas: a most useful type of call

TODO




<!----------------------------------------------------------------------------->

# Environments

Environments are a data structure that are used by R to implement lexical
scoping.

  * When constructing an interpreter, a common approach to scoping is to
    construct a name-value map for each scope, and then traverse the mappings
    until the appropriate match is found or is determined to not exist.
  * R follows this scheme, but goes the additional step of making environments a
    first-class object.
  * Expressions can be evaluated in a chosen environment, which provides a
    mechanism for authors to override the usual scoping rules.




<!----------------------------------------------------------------------------->

# Promises and lazy evaluation

## Lazy evaluation in R

The S language stipulates that the arguments of a function call are not
evaluated until they are needed.

  * A form of _lazy evaluation_.
  * Implemented in R by assigning each formal argument name a special object of
    type _promise_. <!-- Section 13.2, Software for Data Analysis -->
  * As an interesting aside, R package lazy loading of data is implemented on
    top of promises.




## Promises

The _promise_ type has the following characteristics.

  * Contains three elements:
    * The expression provided as input to the argument in the function call.
    * The environment in which to evaluate the expression.
    * The value obtained by evaluating the expression (initially empty).
  * The value of the promise is calculated the first time it is needed, and not
    before.  After calculation the value is stored.
  * Subsequent uses of the variable use the stored value.




## Extracting the promise expression

The `substitute` function captures an expression in a manner similar to `quote`.

  * However, `substitute` performs the added step of replacing symbols that
    it finds in the expression that exist in the specified environment.
  * The environment is specified through the formal argument `env` (the default
    is the current execution environment).
  * If a symbol in the expression has a binding in the specified environment,
    then it replaces the symbol with the value bound to the symbol in the
    specified environment or list object.
  * As a special case, if a given variable name is bound to an unevaluated
    promise, then `substitute` returns the _expression_ associated with the
    promise.

```{r}
# non-promise object
substitute(x + y, list(y = 11 + 22))

# promise object
f <- function(x) substitute(x + y)
f(11 + 22)
```




## Extracting promise expressions in rlang

  * `rlang::enexpr` obtains the promise expression of a given argument.
  * `rlang::enexprs` returns a list with each element the promise expression
    corresponding to an argument passed to `...`.

Base R doesn't have an equivalent to `rlang::enexprs`, but Advanced R suggests
the clever hack `as.list(substitute(...()))`.  Also see `match.call` to obtain
the entire function call.








<!-- promises page 457 Software for Data Analysis ----------------------------->

<!-- R implements the grammer of S, with a few extensions --------------------->

<!-- http://www.sumsar.net/blog/2014/11/tidbits-from-books-that-defined-s-and-r/ -->

<!-- https://www.stat.auckland.ac.nz/~ihaka/downloads/Interface98.pdf -->

<!-- https://rpubs.com/maechler/R_language_objs -->
